Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-09-29 Tex/Paper.tex
Edits before sending to Erik and Patric

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativistic charged particle beam
travelling through a plasma excites a strong longitudinal electric field that can be loaded by a trailing witness beam.
With optimal beam loading, an increase in absolute energy spread can be kept to a minimum. Acceleration of an electron
withness beam by an electron drive beam has been demonstrated experimentally in the past
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008}. AWAKE at CERN is a proof of concept proton driven plasma wakefield
accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the
accelerated beam. The beam will also see a transversely and longitudinally varying accelerating field causing an
increase in energy spread \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by
the transverse oscillations of the plasma electrons forming a sheet around an evacuated area filled with only ions. The
ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly with
radius. This produces a focusing force that preserves emittance \cite{lu:2006-1, lu:2006}.
%~ This effect also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread
%~ is still dependent on beam loading $E_{z}$.
%~ Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
%~ \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where
%~ \begin{align}
    %~ \lambda_{pe} = 2\pi\frac{c}{\omega_{pe}} \label{EQ:LambdaP}
%~ \end{align}
%~ and
%~ \begin{align}
    %~ \omega_{pe} = \sqrt{\frac{n_{0} e^{2}}{m_{e} \epsilon_{0}}}, \label{EQ:OmegaP}
%~ \end{align}
$\lambda_{pe} = 2\pi c/\omega_{pe}$ and $\omega_{pe} = (n_{0} e^{2} / m_{e} \epsilon_{0})^{1/2}$,
produces a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness beam loading the
peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches when they are
properly phased with respect to the wakefields. Acceleration of an electron witness beam driven by two electron drive
beams has been demonstrated at Brookhaven National Laboratory \cite{muggli:2011}.

The energy carried by electron drive bunches used in previous experiments were typically small -- on the order of
$100\unit{J}$ -- and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. The energy of a
high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam that could be produced by the international
linear collider with $1\nexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers pulses as drivers, a
large number of plasma stages is required. Staging plasma accelerators without reducing the effective gradient and
spoiling the beam quality is challenging \cite{steinke:2016,lindstrom:2016}. Proton beams available at CERN carry
significantly more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for driving longer plasma
wakefield accelerator stages. However, the SPS beam is orders of magnitude longer than the plasma wavelengths needed for
such applications. This issue is resolved by letting the proton beam undergo self-modulation before injecting the
witness beam into one of the buckets in the modulated structure. This self-modulation process is caused by the
transverse fields generated by the beam acting upon the beam itself, causing regions of the beam to rapidly defocus
\cite{kumar:2010}. The modulation frequency is close to that of the plasma, and produces a train of short proton bunches
along the beam axis with a surrounding halo of defocused particles. The train resonantly produces wakefields to larger
amplitudes.
%~ De-phasing of the
%~ witness beam relative to the drive beam, caused by loss of energy in the drive beam and gain of energy in the witness
%~ beam, limits the effective length of a plasma accelerator stage.

@@

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver and a nominal plasma density of $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the transverse size
of the SPS proton beam such that $k_{pe}\sigma_{x,y} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number.
The aim of the first phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam. The aim of
the second phase in 2018 is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented
here is for Run 2 \cite{adli:2016}, which aims to demonstrate acceleration of a short electron beam, $\ll\lambda_{pe}$,
to high energy with a minimal increase in emittance and absolute energy spread.

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={1mm 2mm 1mm 2mm},clip]{figures/figAWAKE}
    \caption{\label{Fig:AWAKER2} A simplified illustration of the experimental setup for AWAKE Run 2. The SPS proton
        beam undergoes self-modulation in the first $4\unit{m}$ plasma section. The electron witness beam is injected
        into the accelerating structure, and undergoes acceleration in the second plasma stage
        \cite{berglyd_olsen:2015, adli:2016}.}
\end{figure}

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}. The first
section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the electron
beam present. The electron witness beam is then  injected into the modulated proton beam before stage two, where it
undergoes acceleration.
%~ As the $E_z$-field will decrease due to the gap between the two cells, it is desirable to
%~ keep this as short as possible.

Using the SPS proton beam as driver, the self-modulated proton beam does not produce a full non-linear wakefield, and
not all plasma electrons are therefore evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius. The small size of the accelerating structure produced by the self-modulated beam also
puts constraints on the size of the electron witness beam, and thus its charge and current -- if at the same time we
wish to prevent large energy spread. By matching the transverse size of the beam at a given emittance to the plasma
density, we prevent large amplitude oscillations which may cause additional growth in energy spread as well as in
emittance.

The key idea is to have enough charge in the witness bunch to at the same time load the wakefield to produce low
relative energy spread $\Delta E/E$ and blow out the electrons left in the accelerating structure to reach conditions
that preserves emittance.
%~ The key idea in this paper is that while the proton beam wake is not fully blown-out, and thus leading to non-linear
%~ focusing forces, the electron beam, if intense enough, may load the field and at the same time create its own ion
%~ bubble. As will be shown, exploiting these two well known effects, the electron beam can be accelerated for long
%~ distances without significant emittance growth.

%~ The beam matching relation is given by
%~ \begin{align}
    %~ \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_pe}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
%~ \end{align}
%~ where $\lambda_{p}$ is the plasma wavelength. For a reasonable initial emittance of $2\unit{\mu m}$, this corresponds to
%~ a $\sigma_{x}$ of $5.25\unit{\mu m}$ -- a very narrow bunch compared to the drive beam $\sigma_{x} = 200\unit{\mu m}$.
%~ The charge density of this compact witness bunch reaches that of the plasma density at only a few $\unit{pC}$, but
%~ reaches optimal beam loading at around $1-200\unit{pC}$. The implication here being that the witness bunch produces its
%~ own non-linear wakefield already at the head of the bunch. The majority of the electrons within it will therefore see a
%~ linear focusing force preventing the bulk of the beam from undergoing emittance growth.

@@

The main focus of this study is on the loading of the wakefields driven by the proton beam. In order to eliminate other
factors that may affect this (add examples?), we tried several approaches to create a stable drive beam structure based
on previous self-modulation studies.

Our previous approach was to use a pre-modulated, short proton beam with similar structure to the one produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations \cite{berglyd_olsen:2015, berglyd_olsen:2016}.

@@

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. When
the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second plasma cell, the peak electric
field is expected to be $500-600\funit{MV}{m}$. The plasma electrons are only depleted to around $65\%$ of nominal
plasma density at the point where we inject the electron beam \cite{awake_collaboration:2016}. These conditions are
replicated in our simulations using a single proton beam of $1.46\nexp{10}$ protons, or $2.34\unit{nC}$ and
$7\unit{kA}$.

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$ and transverse size $\sigma_{x} = 200\unit{\mu m}$. We
also set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. We are not considering the evolution of the
proton beam itself in this study, so in order to create a stable environment to study the evolution of the witness beam
we prevent the proton beam from evolving radially by increasing the particle mass 6 orders of magnitude.

@@

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}.
To eliminate the problem of initial de-phasing of the witness beam with respect to the proton wakefield, initial beam
energy was set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. The relative drift causing the de-phasing is proportional
to $\gamma^{-2}$. By matching the initial $\gamma$ values in these simulations we eliminate this effect entirely.
However, a lower initial value is likely to be sufficient \cite{berglyd_olsen:2015}.

In these simulations we have used a matched beam for an initial normalised emittance of $\epsilon_{N} = 2\unit{\mu m}$.
The beam matching relation is given by
\begin{equation}
    \beta = \frac{\sigma_{x,y}^{2}\gamma}{\epsilon_{N}}
          = \frac{c}{\omega_{pe}}\sqrt{2\gamma}, \label{EQ:MatchedB}
    % = c\sqrt{\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}}2\gamma}
\end{equation}
where $\beta$ is the Twiss parameter. It then follows that the transverse profile of the beam must satisfy
\begin{equation}
    \sigma_{x,y}^{2} = c\epsilon_{N}\sqrt{2\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}} \label{EQ:MatchedR}
\end{equation}
in order to be matched to a given plasma density, initial emittance and energy. This relation requires a very narrow
beam with a $\sigma_{x,y}$ of $5.25\unit{\mu m}$ compared to the drive beam $\sigma_{x,y} = 200\unit{\mu m}$. The charge
density of this compact witness beam reaches that of the plasma density at only a few $\unit{pC}$, but reaches optimal
beam loading at around $1-200\unit{pC}$. The implication here being that the witness beam produces its own non-linear
wakefield already at the head of the beam. The majority of the electrons within it will therefore see a linear focusing
force preventing the bulk of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam as we need a simulation box wide enough to contain the larger beam while resolving the
narrower one. We used a transverse grid cell size of $1.17\unit{\mu m}$, and of $2.34\unit{\mu m}$ for the longitudinal
grid cells for the simulations presented in section \ref{S:BL}. The witness beam was simulated with $16.8\unit{M}$
non-weighted particles. The witness beam has a total charge of $100\unit{pC}$, a $\sigma_{z}=60\unit{\mu m}$, and a
$\sigma_{x}=5.25\unit{\mu m}$ matching an initial normalised emittance $\epsilon_{0} = 2\unit{\mu m}$. We refer to this
parameter set as our base case.

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} The top plot shows the unloaded longitudinal electric field with no witness beam
        (dashed blue line) and the loaded field (whole blue line) along the beam axis. The beam density along the axis
        for both beams is shown in red line. The bottom plot compares the plasma density along the beam axis for a drive
        beam with no witness beam (dashed green line), witness beam with no drive beam (dash-dotted green line), and
        both beams present (whole green line). $\xi = z - tc$ is the position in the simulation box, travelling towards
        the left. the plots are normalised to $n_{0} = 7\nexp{-14}\unit{cm}^{-3}$.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamPhaseSpace}
    \caption{\label{Fig:BeamPS} Phase space charge distribution of a $100\unit{pC}$, $60\unit{\mu m}$ long witness beam
        after $4\unit{m}$ of plasma. The mean momentum is $1.67\unit{GeV/c}$ with an RMS of $87\unit{MeV/c}$ ($5.2\%$){}
        for the full beam.}
\end{figure}

\begin{figure*}[hbt]
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittance}
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittanceOffset}
    \caption{\label{Fig:BeamEmitt} The red lines show a moving window calculation of transverse normalised emittance for
        an on-axis beam with respect to the drive beam axis (left), and an offset beam (right) with an offset of one
        $\sigma_{x} = 5.24\unit{\mu m}$ in the x-plane. The moving window for emittance calculation is longitudinal
        slices of $l = 4\times\Delta\xi = 9.38\unit{\mu m}$ with a $\Delta\xi$ resolution. Only slices with more than
        $100$ macro particles have been included. The blue lines show the on-axis electron beam density profile. For
        reference, the plasma density profile is included in green, but scaled up by a factor of $100$ to be visible.
        The solid lines are taken at the plasma entry point, the dashed lines after $4\unit{m}$ of plasma, the
        dash-dotted after $40\unit{m}$, and the dotted after $100\unit{m}$. These simulations were run with an LHC
        energy drive beam of $7\unit{TeV}$}
\end{figure*}

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    %~ \includegraphics[height=22.2mm,trim={ 0mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation01}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  6mm 0mm},clip]{figures/beamFilamentation02}
    %~ \includegraphics[height=22.2mm,trim={16mm 9.5mm  0mm 0mm},clip]{figures/beamFilamentation03}
    %~ \includegraphics[height=22.2mm,trim={10mm 9.5mm 15mm 0mm},clip]{figures/beamOffsetMean}
    %~ \\
    %~ \includegraphics[height=26.0mm,trim={ 0mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation04}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  6mm 2mm},clip]{figures/beamFilamentation05}
    %~ \includegraphics[height=26.0mm,trim={16mm 0.1mm  0mm 2mm},clip]{figures/beamFilamentation06}
    %~ \includegraphics[height=26.0mm,trim={10mm 0.1mm 15mm 2mm},clip]{figures/beamOffsetRMS}
    \caption{\label{Fig:BeamFilament} The left plots show the transverse phase space of the electron beam at different
        plasma positions. The right side shows the macro particle mean position (top) and RMS spread (bottom). The blue
        particles and lines represent particles with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. The red
        particles and lines represent particles with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

% ******************************************************************************************************************** %
\section{Beam Loading}\label{S:BL}
%~ Eric's Comments
%~ Sec N: Beam loading [in this section we describe the interesting physics, based on a single parameters set]
%~
%~ Patric's Comments
%~ - Beam loading for narrow energy spread
%~ - Beam loading in linear regime (old papers from SLAC in the 80's, then Katsouleas for PWFA) and nonlinear regime
%~   (Tzoufras) well know
%~ - Propose "new regime" in the SM case, beam loading and blowout to get emittance preservation
%~ - Limitation as for PWFA driver, must use some of the W bunch to reach the blowout, it is kind of like the early SLAC
%~   experiments (84GeV) where the W-bunch is like the D-bunch in these experiments, but in a plasma prepared by the p+
%~   bunch and the SM
%~ - maybe also something about the fact that 200pC in a 60µm bunch corresponds to a 1kA current, much larger than the
%~   p+ bunch current, because of the fact that the wakefields are driven my multiple bunches?
% ******************************************************************************************************************** %

%~ - describe proton beam wake structure [similar to that of a modulated proton beam] (simulation results to show:
%~   unloaded wake)

@@

%~ - by varying the current of the electron beam, the electron beam may load the wake and the longitudinal field can be
%~   flatten and energy spread reduced [well known result, Tzoufras]

@@

%~ - the acceleration and the transverse emittance stable on long timescales (simulation results to show: transverse and
%~   longitudinal phase spaces, space long simulation)

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
For the majority of the cases we studied that maintained a stable accelerating structure, about $70-80\%$ of the
electron beam retained its initial emittance. Figure \ref{Fig:BeamEmitt} shows emittance growth for the base case as a
function of propagation length. No significant emittance growth was observed for propagation lengths up to
$100\unit{m}$. The drive beam energy was here increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing, as
de-phasing starts to become a significant effect for the SPS beam of $400\unit{GeV}$ after about $50\unit{m}$.

>>>>

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. With a relativistic charged particle drive
beam travelling through a plasma, a strong wakefield is excited, that can be loaded by a trailing witness beam. When the
witness beam optimally loads the wakefield, an increase in absolute energy spread can be kept to a minimum. The concept
has been demonstrated experimentally in the past using electron drive beams accelerating electron witness beams
\cite{rosenzweig:1988, blumenfeld:2007, kallos:2008}. AWAKE at CERN is a proof of concept beam driven plasma wakefield
accelerator experiment using a proton drive beam delivered by the SPS accelerator \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to accelerate a beam while keeping energy spread and
emittance growth to a minimum. In the well described linear case where the beam density $n_{b}$ is much smaller than the
plasma density $n_{0}$, a non-linear transverse focusing force causes emittance growth of the witness beam. the beam
will also see a transversely and longitudinally varying accelerating field causing a spread in energy after the beam has
been accelerated \cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by the
transverse oscillations of the plasma electrons, forming a sheet around an evacuated area filled with only ions. The
ions, assumed stationary, form an uniform density ion channel creating a focusing force that varies linearly with
radius. This produces a focusing force that preserves emittance \cite{lu:2006-1, lu:2006}.

In this paper we present simulation results showing how we can use the non-linear wakefields driven by the head of a
high charge density witness beam to reduce energy spread and emittance growth in the rest of the beam, while being
accelerated by a proton drive beam producing quasi-linear wakefields.

@@

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where the
plasma wavelength $\lambda_{pe} = 2\pi c/\omega_{pe}$ and the plasma frequency $\omega_{pe} = (n_{0} e^{2} / m_{e}
\epsilon_{0})^{1/2}$, produces a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness
beam loading the peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches
when they are properly phased with respect to the wakefields. Acceleration of an electron witness beam driven by two
electron drive beams has been demonstrated at Brookhaven National Laboratory \cite{muggli:2011}.

The energy carried by electron drive bunches used in previous experiments were typically small -- on the order of
$100\unit{J}$ -- and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. Higher energy
electron beam can be improve the propagation length and thus the energy of the accelerated beam. For instance, the
energy of a high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam that could be produced by the
international linear collider with $1\nexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers pulses
as drivers, a large number of plasma stages is required. However, staging plasma accelerators without reducing the
effective gradient and spoiling the beam quality is challenging \cite{steinke:2016,lindstrom:2016}.

Proton beams available at CERN carry significantly more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016},
allowing for much longer plasma wakefield accelerator stages. The SPS beam is, unfortunately, orders of magnitude longer
than the plasma wavelengths needed for such applications. This can be resolved by letting the proton beam undergo
self-modulation before injecting the witness beam into accelerating structure. The self-modulation of the beam is
produced by the transverse fields generated by the beam acting upon itself, causing regions of the beam to rapidly
defocus \cite{kumar:2010}. The modulation frequency is close to that of the plasma, and produces a train of short proton
bunches along the beam axis with a surrounding halo of defocused particles. This train of bunches resonantly produces
wakefields to larger amplitudes.

@@

The AWAKE experiment, currently in its first stages of operation at CERN, uses a $400\unit{GeV}$ beam delivered by the
SPS as its driver, and a single $10\unit{m}$ plasma stage with a nominal plasma density of $7\nexp{14}\unit{cm}^{-3}$
\cite{gschwendtner:2016}. This plasma density corresponds to $\lambda_{pe} = 1.26\unit{mm}$ -- matched to the transverse
size of the SPS proton beam such that $k_{pe}\sigma_{x,y} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave
number.

The aim of the first phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam. The aim of
the second phase in 2018 is to sample the wakefield with a long electron beam $\simeq\lambda_{pe}$. The study presented
here is for Run 2 \cite{adli:2016}, which aims to demonstrate acceleration of a short electron beam, $\ll\lambda_{pe}$,
to high energy with a minimal increase in emittance and absolute energy spread.

\begin{figure}[hbt]
    \includegraphics[width=0.99\linewidth,trim={1mm 2mm 1mm 2mm},clip]{figures/figAWAKE}
    \caption{\label{Fig:AWAKER2} A simplified illustration of the experimental setup for AWAKE Run 2. The SPS proton
        beam undergoes self-modulation in the first plasma section. The electron witness beam is injected into the
        accelerating structure, and undergoes acceleration in the second plasma stage
        \cite{berglyd_olsen:2015, adli:2016}.}
\end{figure}

The plans for AWAKE Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}. The first
section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without the electron
beam present. The electron witness beam is then injected into the modulated proton beam before stage two, where it
undergoes acceleration.

Using the SPS proton beam as driver, the self-modulated proton beam does not produce a full non-linear wakefield, and
not all plasma electrons are therefore evacuated from the plasma bubble. The result is that the focusing force does not
increase linearly with radius. The small size of the accelerating structure produced by the self-modulated beam also
puts constraints on the size of the electron witness beam, and thus its charge and current, if at the same time we wish
to prevent large energy spread. By matching the transverse size of the beam at a given emittance to the plasma density,
we prevent large amplitude oscillations which may cause additional energy spread as well as emittance growth.

%~ Eric's version:
%~ The key idea in this paper is that while the proton beam wake is not fully blown-out, and thus leading to non-linear
%~ focusing forces, the electron beam, if intense enough, may load the field and at the same time create its own ion
%~ bubble. As will be shown, exploiting these two well known effects, the electron beam can be accelerated for long
%~ distances without significant emittance growth.

%~ Patric's version:
The key idea is to have enough charge in the witness beam to at the same time load the wakefield to produce low
relative energy spread $\Delta E/E$, and blow out the electrons left in the accelerating structure to reach conditions
that preserves emittance.

@@

A main focus of this study is on the loading of the wakefields driven by the proton beam. In order to eliminate other
factors that may affect this, like decay and defocusing of the proton beam, we tried several approaches to create a
stable drive beam structure based in previous self-modulation studies.

In these studies we used a short, pre-modulated proton beam with similar structure to the beam produced by the
self-modulation of the SPS beam in AWAKE. These studies were done using the full PIC code Osiris \cite{fonseca:2002}
using 2D cylindrical-symmetric simulations and primarily looked at beam loading, energy gain and energy spread
\cite{berglyd_olsen:2015, berglyd_olsen:2016}.

@@

The baseline AWAKE drive beam current is insufficient to reach the non-linear regime and produce a plasma bubble. When
the SPS beam, containing $3\nexp{11}$ protons \cite{gschwendtner:2016}, enters the second plasma stage, the peak
electric field is expected to be $500-600\funit{MV}{m}$. The plasma electrons are only depleted to around $65\%$ of
nominal plasma density at the point where we inject the electron beam \cite{awake_collaboration:2016}. In our simulation
these conditions are replicated using a single proton beam of $1.46\nexp{10}$ protons, or $2.34\unit{nC}$ and
$7\unit{kA}$.

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton beam energy of $400\unit{GeV}$ and transverse size $\sigma_{x} = 200\unit{\mu m}$. We
also set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. We are not considering the evolution of the
proton beam itself in this study, so in order to create a stable environment to study the evolution of the witness beam
we prevent the proton beam from evolving radially by increasing the particle mass by 6 orders of magnitude.

@@

Beam loading of a short witness beam is sensitive to its position relative to the electric field \cite{tzoufras:2009}.
A low gamma witness beam can easily drift out of phase with the drive beam wakefield. The drift of the beams is
proportional to $\gamma^{-2}$, so to eliminate initial de-phasing of the witness beam, initial beam energy was set such
that $\gamma_{eb} = \gamma_{pb} = 426.3$, giving an energy of $217\unit{MeV}$. A lower initial value is likely to be
sufficient \cite{berglyd_olsen:2015}.

In these simulations we have used a matched beam for an initial normalised emittance of $\epsilon_{N} = 2\unit{\mu m}$.
The beam matching relation is given by
\begin{equation}
    \beta = \frac{\sigma_{x,y}^{2}\gamma}{\epsilon_{N}}
          = \frac{c}{\omega_{pe}}\sqrt{2\gamma}, \label{EQ:MatchedB}
    % = c\sqrt{\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}}2\gamma}
\end{equation}
where $\beta$ is the Twiss parameter. It then follows that the transverse profile of the beam must satisfy
\begin{equation}
    \sigma_{x,y}^{2} = c\epsilon_{N}\sqrt{2\frac{m_{e}\varepsilon_{0}}{n_{pe}e^{2}\gamma}} \label{EQ:MatchedR}
\end{equation}
in order to be matched to a given plasma density, initial emittance and energy. This relation requires a very narrow
beam with a $\sigma_{x,y}$ of $5.25\unit{\mu m}$ compared to the drive beam $\sigma_{x,y} = 200\unit{\mu m}$. The charge
density of this compact witness beam reaches that of the plasma density at only a few $\unit{pC}$, but reaches optimal
beam loading at around $100-200\unit{pC}$. The implication here being that the witness beam produces its own non-linear
wakefield already at the head of the beam. The majority of the electrons within it will therefore see a linear focusing
force preventing the bulk of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam as we need a simulation box wide enough to contain the larger beam while resolving the
narrower one (see figure \ref{Fig:PlasmaDenTWake}). We used a transverse grid cell size of $1.17\unit{\mu m}$, and of
$2.34\unit{\mu m}$ for the longitudinal grid cells for the simulations presented in section \ref{S:BL}. The witness beam
was simulated with $16.8\nexp{6}$ non-weighted particles. The witness beam has a total charge of $100\unit{pC}$, a
$\sigma_{z}=60\unit{\mu m}$, and a $\sigma_{x}=5.25\unit{\mu m}$ matching an initial normalised emittance
$\epsilon_{0} = 2\unit{\mu m}$. We refer to this parameter set as our base case.

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} The top plot shows the unloaded longitudinal electric field with no witness beam
        (dashed blue line) and the loaded field (whole blue line) along the beam axis. The beam density along the axis
        for both beams are shown in red. The bottom plot compares the plasma densities along the beam axis for a drive
        beam with no witness beam (dashed green line), witness beam with no drive beam (dash-dotted green line), and
        both beams present (whole green line). $\xi = z - tc$ is the position in the simulation box, moving towards the
        left.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamPhaseSpace}
    \caption{\label{Fig:BeamPS} The phase space charge distribution of a $100\unit{pC}$, $60\unit{\mu m}$ long witness
        beam after $4\unit{m}$ of plasma. The mean momentum is $1.67\unit{GeV/c}$ with an RMS energy spread of
        $87\unit{MeV/c}$ ($5.2\%$) for the full beam.}
\end{figure}

\begin{figure*}[hbt]
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittance}
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittanceOffset}
    \caption{\label{Fig:BeamEmitt} Beam density in blue along the beam axis for an on-axis beam with respect to the
        drive beam axis (left), and an offset beam (right) with an offset of one $\sigma_{x} = 5.24\unit{\mu m}$ in the
        x-plane. At four different positions $z$ in the plasma stage. The red lines show a moving window calculation of
        transverse normalised emittance. The moving window calculation uses longitudinal slices of
        $l = 4\times\Delta\xi = 9.38\unit{\mu m}$ with a step of $\Delta\xi$. Only slices with more than $100$ macro
        particles have been included. The plasma density profile is included in green, and scaled up by a factor of
        $100$ to be visible. These simulations were run with an LHC energy drive beam of $7\unit{TeV}$}
\end{figure*}

\begin{figure*}[hbt]
    \includegraphics[width=\linewidth,trim={0mm 0mm 0mm 0mm},clip]{figures/beamFilamentationAll}
    \caption{\label{Fig:BeamFilament} Plots \textbf{a} to \textbf{f} show the transverse phase space of the offset
        electron beam at different plasma positions. Plot \textbf{g} shows the macro particle mean position, and plot
        \textbf{h} their RMS spread. Plots \textbf{a}, \textbf{b} and \textbf{c}, as well as the blue lines in plots
        \textbf{g} and \textbf{h} represent particles with position $1.40\unit{\mu m} < \xi < 1.42\unit{\mu m}$. Plots
        \textbf{d}, \textbf{e} and \textbf{f}, and the red lines in plots \textbf{g} and \textbf{h} represent particles
        with position $1.55\unit{\mu m} < \xi < 1.57\unit{\mu m}$.}
\end{figure*}

% ******************************************************************************************************************** %
\section{Beam Loading}\label{S:BL}
%~ Eric's Comments
%~ Sec N: Beam loading [in this section we describe the interesting physics, based on a single parameters set]
%~ - describe proton beam wake structure [similar to that of a modulated proton beam] (simulation results to show:
%~   unloaded wake)
%~ - by varying the current of the electron beam, the electron beam may load the wake and the longitudinal field can be
%~   flatten and energy spread reduced [well known result, Tzoufras]
%~ - the electron beam blows out the remaining plasma electrons. The transverse fields are dominated by the linear
%~   focusing fields originating from the ion background, leading to emittance preservation of the part of the electron
%~   beam inside the bubble [loading of a quasi-linear wake, and still get emittance preservation: this is new]
%~   (simulation results to show: loaded wake, with bubble - many different aspects of this, trans., long. field,
%~   electron densities etc. )
%~ - the acceleration and the transverse emittance stable on long timescales (simulation results to show: transverse and
%~   longitudinal phase spaces, space long simulation)
%~ - in this regime, emittance preservation is robustness to drive-witness offset [this is new] (simulation results to
%~   show: transverse and longitudinal phase spaces, long offset simulation)
%~
%~ Patric's Comments
%~ - Beam loading for narrow energy spread
%~ - Beam loading in linear regime (old papers from SLAC in the 80's, then Katsouleas for PWFA) and nonlinear regime
%~   (Tzoufras) well know
%~ - Propose "new regime" in the SM case, beam loading and blowout to get emittance preservation
%~ - Limitation as for PWFA driver, must use some of the W bunch to reach the blowout, it is kind of like the early SLAC
%~   experiments (84GeV) where the W-bunch is like the D-bunch in these experiments, but in a plasma prepared by the p+
%~   bunch and the SM
%~ - maybe also something about the fact that 200pC in a 60µm bunch corresponds to a 1kA current, much larger than the
%~   p+ bunch current, because of the fact that the wakefields are driven my multiple bunches?
% ******************************************************************************************************************** %

@@

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
For the majority of the cases we studied, that maintained a stable accelerating structure, about $70-80\%$ of the
electron beam retained its initial emittance. Figure \ref{Fig:BeamEmitt} shows emittance along the beam for the base
case, sampled after propagating through $0$, $4$, $40$ and $100\unit{m}$ of plasma. Emittance growth mainly occurs in
first few meteres, and no significant emittance growth was observed after this for propagation lengths up to
$100\unit{m}$. For this simulation, drive beam energy was increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing,
as de-phasing starts to become a significant effect for the SPS beam of $400\unit{GeV}$ after about $50\unit{m}$.
